---
title: "Explaining the Bithday Problem"
author: "Fernando Hoces de la Guardia"
date: "09/03/2021"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
n.pers = 21
n.sims = 1e4
```

# Birthday problem  
What is the probability that at least two people share the same birhtday in this room?

Is is $\frac{1}{365} \times `r n.pers` = `r 1/365 * n.pers`$


## Analytical solution  

But actually when we compute the math. We get  an  surprising result: 

\begin{align}   
 1 - \bar p(n) &= 1 \times \left(1-\frac{1}{365}\right) \times \left(1-\frac{2}{365}\right) \times \cdots \times \left(1-\frac{n-1}{365}\right) \nonumber  \\  &= \frac{ 365 \times 364 \times \cdots \times (365-n+1) }{ 365^n } \nonumber \\ &= \frac{ 365! }{ 365^n (365-n)!} = \frac{n!\cdot\binom{365}{n}}{365^n}\\
p(n= `r n.pers`) &= `r  round(1 - factorial(n.pers) * choose(365,n.pers)/ 365^n.pers, 3)`  \nonumber
\end{align}


## Simulations  

1 - Simulate 10,000 rooms with n = `r n.pers` random birthdays, and store the results in matrix where each row represents a room.  
2 - For each room (row) compute the number of unique birthdays.  
3 - Compute the average number of times a room has `r n.pers` unique birthdays, across 10,000 simulations, and report the complement.  

  
```{r}
birthday.prob = function(n.pers_var, n.sims_var) {
  # simulate birthdays
  birthdays = matrix(round(runif(n = n.pers_var * n.sims_var, min = 1, max = 365) ),
                     nrow = n.sims_var, ncol = n.pers_var)
  # for each room (row) get unique birthdays
  unique.birthdays = apply(birthdays, 1,
                           function(x)  length( unique(x) ) )
  # Indicator with 1 if all are unique birthdays
  all.different = 1 * (unique.birthdays==n.pers_var)
  # Compute average time all have different birthdays 
  result = 1 - mean(all.different)
return(result)
}

bp_sim = birthday.prob(n.pers_var = 21, n.sims_var = 10000)
print(bp_sim)
```

## Results 

- Many people originally think of a prob ~ $\frac{1}{365} \times n =   `r round(1/365 * n.pers, 3)`$

- However the true probability is of $p(n= `r n.pers`) = `r round(1 - factorial(n.pers) * choose(365,n.pers)/ 365^n.pers, 3)`$

- And the simulated probability is of  $`r bp_sim`$
