---
title: "First Slides: Birthday Problem"
author: "Fernando Hoces de la Guardia"
date: "9/1/2021"
output:
  beamer_presentation: default
  slidy_presentation: default
  ioslides_presentation: default
---

<!--
- for HTML, slidy seems to be the most compatible,
- for all: replace non r code chunk headers with "md" (compare line 42 with 61 )
- for beamer, replace \newline with "\\"
- general point: commit to one ouput in slides, don't try to switch back and forth
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
n.pers = 21
n.sims = 1e4
```


## The birthday problem: the math
Actually the math says otherwise. Define p(n) as the probability that at least one pair has the same birthday, then the $1 - p(n)$ is the probability that all are born in a different day. Which we can compute as: 


\begin{align} 
 1 -  p(n) &= 1 \times \left(1-\frac{1}{365}\right) \times \left(1-\frac{2}{365}\right) \times \cdots \times \left(1-\frac{n-1}{365}\right) \nonumber  \newline
 &= \frac{ 365 \times 364 \times \cdots \times (365-n+1) }{ 365^n } \nonumber \newline
 &= \frac{ 365! }{ 365^n (365-n)!} = \frac{n!\cdot\binom{365}{n}}{365^n}\newline
p(n= `r n.pers`) &= `r  round(1 - factorial(n.pers) * choose(365,n.pers)/ 365^n.pers, 3)`  \nonumber
\end{align}


## Code for the math 

(`/dynamicdocs/first_dd_solution.Rmd`)

Copy and paste lines below into your `first_dd.Rmd`

```md
\begin{align} 
 1 -  p(n) &= 1 \times \left(1-\frac{1}{365}\right) \times
              \left(1-\frac{2}{365}\right) \times \cdots \times
              \left(1-\frac{n-1}{365}\right) \nonumber  \newline
           &= \frac{365 \times 364 \times \cdots \times (365-n+1) }{365^n} \nonumber \newline
           &= \frac{ 365! }{ 365^n (365-n)!} = \frac{n!\cdot\binom{365}{n}}{365^n}\newline
\p(n= `r n.pers`) &= `r  round(1 - factorial(n.pers) * 
                          choose(365,n.pers)/ 365^n.pers, 3)`  \nonumber
\end{align}
```


## Code for the math 

(`/dynamicdocs/first_dd_solution.Rmd`)

Copy and paste lines below into your `first_dd.Rmd`

```{latex, eval=FALSE}
\begin{align} 
 1 -  p(n) &= 1 \times \left(1-\frac{1}{365}\right) \times
              \left(1-\frac{2}{365}\right) \times \cdots \times
              \left(1-\frac{n-1}{365}\right) \nonumber  \newline
           &= \frac{365 \times 364 \times \cdots \times (365-n+1) }{365^n} \nonumber \newline
           &= \frac{ 365! }{ 365^n (365-n)!} = \frac{n!\cdot\binom{365}{n}}{365^n}\newline
*\p(n= `r n.pers`) &= `r  round(1 - factorial(n.pers) * 
*                          choose(365,n.pers)/ 365^n.pers, 3)`  \nonumber
\end{align}
```


## Don't like math? Let's run a simple simulation!   


 1 - Simulate `r n.sims` rooms with $n = `r n.pers`$ random birthdays, and store the results in matrix where each row represents a room.   
 
 2 - For each room (row) compute the number of unique birthdays.   
 
 3 - Compute the average number of times a room has `r n.pers` unique birthdays, across `r n.sims` simulations, and report the complement.    


## Code for the simulation 

(`/dynamicdocs/first_dd_solution.Rmd`)  


```{r birthday prob, eval=TRUE, echo=TRUE}
birthday.prob = function(n.pers_var, n.sims_var) {
  # simulate birthdays
  birthdays = matrix(round(runif(n = n.pers_var * n.sims_var, min = 1, max = 365) ),
                     nrow = n.sims_var, ncol = n.pers_var)
  # for each room (row) get unique birthdays
  unique.birthdays = apply(birthdays, 1,
                           function(x)  length( unique(x) ) )
  # Indicator with 1 if all are unique birthdays
  all.different = 1 * (unique.birthdays==n.pers_var)
  # Compute average time all have different birthdays 
  result = 1 - mean(all.different)
return(result)
}

bp_sim = birthday.prob(n.pers_var = 21, n.sims_var = 10000)
print(bp_sim)
```



## Results   


- Many people originally think of a prob ~ $\frac{1}{365} \times n =   `r round(1/365 * n.pers, 3)`$

- However the true probability is of $p(n= `r n.pers`) = `r round(1 - factorial(n.pers) * choose(365,n.pers)/ 365^n.pers, 3)`$

- And the simulated probability is of  $`r bp_sim`$


